Predicting the Quality of CDRH3 Antibody Loop Structural Models

### 2022-10-05 paper.tex
Fixed all full-stops commas and dashes

Several different approaches for generating three-dimensional
structure models from antibody sequences exist such as
RosettaAntibody\cite{Sircar2009,Sivasubramanian2009}, 
ABodyBuilder\cite{Leem2016}, PIGSPro\cite{Lepore2017} and abYmod,
UCL's in-house software developed by Prof. Martin. One of the most
used methods is RosettaAntibody, which implements template selection
and ab initio CDR-H3 loop modelling using loop fragments and employing
specific angle restraints which bias the conformational space towards
so-called `kinked' loops\cite{Schoeder2021,Weitzner2017}. In
contrast, ABodyBuilder uses a database search algorithm
(FREAD\cite{Choi2010}) for CDR loop modelling.

@@

The predictive power of any machine learning model is largely
dependent on the quality and size of the dataset it was trained on. As
this is a non-linear, complex, multi-class classification problem, a
substantial amount of data was required. Thus, an extensive, verified
dataset of antibody structures called abYbank/AbDb\cite{XXXX},
established by Prof. Andrew Martin, was utilised (1924 non-redundant
structures). The root-mean-square deviation (RMSD) value, a measure of
distance between backbone \ca\ atoms of superimposed crystal structures
and modelled structures, is calculated (see methods). This metric for
model quality was used to classify models.

@@

The full pipeline for creating the final machine learning model that
will predict model quality by giving its RMSD range starts with a
feature-set calculation using the antibody sequence. The feature set
includes many different attributes linked to sequence, physical
characteristics, interactions, etc. within as well as outside of the
loop. There is a plethora of information to train our classifier on,
including packing quality, protrusion, hydrophobicity, pseudo-energy
(see protrusion, hydrophobicity, pseudo-energy (see methods for
details). The sequence logo (Figure~\ref{fig:logo}) visualizes amino acid
occurrence within the loop sequence, elements of which can be
extracted as features \cite{Thomsen2012,Shaner1993}.

\begin{figure}
  \includegraphics[scale=0.9]{logo.eps}
  \caption {Sequence Logo of the CDR-H3 loop sequence. Data on amino
    acid occurrence taken from http://abymod.abysis.org/. Visualized
    using Seq2Logo.}
  \label{fig:logo}
\end{figure}

After creating the feature dataset, it is pre-processed (cleaning,
scaling, encoding, see methods for details). Structures with a
resolution below 4\AA\ were removed given their low quality. Identical
whole antibody structures were removed from the dataset, while
instances of different antibodies that matched in loop sequence were
not removed. Models of some of these structures with the same loop
sequence differ significantly. The few large RMSD ranges may stem from
low resolution, e.g.\ the highest datapoint contains a structure with a
resolution of 3.00 \AA.  Residue differences near the loop may also
explain the conformational difference. Some of these structures are
complexed while others are not, which may also affect the loop
structure.

@@

The target data (i.e.\ RMSD values) are then transformed from numerical
values to nominal values so that they can be used for
classification. In order to define these nominal categories, the total
RMSD range must be divided into categories. This is done either by
creating uniform classes i.e.\ 1-2\AA, 2-3\AA\ (the optimal size of which
must be determined), etc. or by creating balanced classes. When
creating balanced classes, the upper and lower thresholds of a
category are chosen in such a way that each class contains an equal
number of instances. This approach is chosen to counteract the
skewness of the RMSD distribution. However, this was found to
negatively affect the final model's predictive power. Therefore,
uniform classes were used.

They are also transformed into a set of binary values according to a
list of RMSD thresholds. This is done so that binary models can be
trained, which will predict the probability e.g.\ that the model's RMSD
is above 2\AA, 2.2\AA, 2.4\AA, and so on. The number of binary
classifiers incorporated into the first layer have a great effect on
the final model, the general trend being that the more binary
classifiers are used, the better the nominal prediction.

@@

The classifier predicts whether a model has an RMSD of below 2\AA,
between 2-4\AA, or above 4\AA. These cut-off values were selected
based on the observation that abYmod generally produces a model with
RMSD below 4\AA. Incorrectly modelled structures (Figure~\ref{fig:AMA}) may
be identified by screening for structures estimated to have an RMSD
above 4\AA. If a very high-quality model is needed one should also
exclude models above 2\AA.

@@

\subsection{Computing}
All machine learning and, feature selection and hyperparameter
optimization algorithms were implemented in Python. The Scikit-learn
library was used for training models, the
Yellowbrick\cite{Bengfort2021} library was utilized for
visualization. All code is available at
https://github.com/LilianDenzler/qualiloop .

@@

\subsubsection{Physiochemical Feature Encoding}
In a paper by L. Nanni and A.Lumini a new encoding technique is
presented which was developed for machine learning classifiers. Many
physiochemical properties are calculated and transformed using a
non-linear Fisher transform for dimensionality reduction.  A vector of
length 19 is produced for each amino acid34.  In a paper on designing
a neural network for predicting the packing angle of the light and
heavy variable chain of an antibodies, A.C.R. Martin and
K.R. Abhinandan introduce an encoding method that produces a
four-dimensional physiochemical feature vector25. The amino acid
properties used are 1) the total number of side-chain atoms, 2) the
number of side-chain atoms in the shortest path from \ca\ to the most
distal atom, 3) the Eisenberg consensus
hydrophobicity\cite{Eisenberg1982}, 4) the charge.

@@

\subsection{Model Assessment} 
Model assessment must be considered at two levels as performance
metrics of binary and multi-class classifiers are calculated
differently and must thus be considered separately. The Mathews
Correlation Coefficient (MCC)\cite{Matthews1975} is deemed the most
informative, taking the ratios of the four confusion matrix categories
into account 39and is thus more reliable than the F1 score and
accuracy . It is also consistent for both binary and multi-class
problems and therefore well suited for our purpose.

@@

\begin{figure}
  \includegraphics[scale=0.6]{angle.eps}
  \caption {Diagram visualizing the process underlying the protrusion
    calculation. First, the base residues (i.e H95 and H102, shown as
    red spheres) of the CDR-H3 (shown in red) are identified. Then, a
    line is drawn between the two \ca\ atoms of these residues (yellow
    dashed line). The distance of the \ca-atom of each residue in the
    CDR-H3 loop to this line is calculated. The residue which has the
    greatest distance to the line (shown as green sphere) is outputted
    as one-letter amino acid code and used as feature. The distance in
    \AA\ (depicted as yellow arrow) is used as the `protrusion'
    feature. }
  \label{fig:angle}
\end{figure}

@@

Given that abYbank/AbDb is soon to be expanded by an additional
ca. 2000 structures, classifier performance on models without a known
abYmod-template sequence may be improved by a larger dataset.

>>>>

Several different approaches for generating three-dimensional
structure models from antibody sequences exist such as
RosettaAntibody\cite{Sircar2009,Sivasubramanian2009}, 
ABodyBuilder\cite{Leem2016}, PIGSPro\cite{Lepore2017} and abYmod,
UCL's in-house software developed by Prof.\ Martin. One of the most
used methods is RosettaAntibody, which implements template selection
and ab initio CDR-H3 loop modelling using loop fragments and employing
specific angle restraints which bias the conformational space towards
so-called `kinked' loops\cite{Schoeder2021,Weitzner2017}. In
contrast, ABodyBuilder uses a database search algorithm
(FREAD\cite{Choi2010}) for CDR loop modelling.

@@

The predictive power of any machine learning model is largely
dependent on the quality and size of the dataset it was trained on. As
this is a non-linear, complex, multi-class classification problem, a
substantial amount of data was required. Thus, an extensive, verified
dataset of antibody structures called abYbank/AbDb\cite{XXXX},
established by Prof.\ Andrew Martin, was utilised (1924 non-redundant
structures). The root-mean-square deviation (RMSD) value, a measure of
distance between backbone \ca\ atoms of superimposed crystal structures
and modelled structures, is calculated (see methods). This metric for
model quality was used to classify models.

@@

The full pipeline for creating the final machine learning model that
will predict model quality by giving its RMSD range starts with a
feature-set calculation using the antibody sequence. The feature set
includes many different attributes linked to sequence, physical
characteristics, interactions, etc., within as well as outside of the
loop. There is a plethora of information to train our classifier on,
including packing quality, protrusion, hydrophobicity, pseudo-energy
(see protrusion, hydrophobicity, pseudo-energy (see methods for
details). The sequence logo (Figure~\ref{fig:logo}) visualizes amino acid
occurrence within the loop sequence, elements of which can be
extracted as features \cite{Thomsen2012,Shaner1993}.

\begin{figure}
  \includegraphics[scale=0.9]{logo.eps}
  \caption {Sequence Logo of the CDR-H3 loop sequence. Data on amino
    acid occurrence taken from http://abymod.abysis.org/ Visualized
    using Seq2Logo.}
  \label{fig:logo}
\end{figure}

After creating the feature dataset, it is pre-processed (cleaning,
scaling, encoding, see methods for details). Structures with a
resolution below 4\AA\ were removed given their low quality. Identical
whole antibody structures were removed from the dataset, while
instances of different antibodies that matched in loop sequence were
not removed. Models of some of these structures with the same loop
sequence differ significantly. The few large RMSD ranges may stem from
low resolution, e.g.\ the highest datapoint contains a structure with a
resolution of 3.00~\AA.  Residue differences near the loop may also
explain the conformational difference. Some of these structures are
complexed while others are not, which may also affect the loop
structure.

@@

The target data (i.e.\ RMSD values) are then transformed from numerical
values to nominal values so that they can be used for
classification. In order to define these nominal categories, the total
RMSD range must be divided into categories. This is done either by
creating uniform classes i.e.\ 1-2~\AA, 2-3~\AA\ (the optimal size of which
must be determined), etc. or by creating balanced classes. When
creating balanced classes, the upper and lower thresholds of a
category are chosen in such a way that each class contains an equal
number of instances. This approach is chosen to counteract the
skewness of the RMSD distribution. However, this was found to
negatively affect the final model's predictive power. Therefore,
uniform classes were used.

They are also transformed into a set of binary values according to a
list of RMSD thresholds. This is done so that binary models can be
trained, which will predict the probability e.g.\ that the model's RMSD
is above 2~\AA, 2.2~\AA, 2.4~\AA, and so on. The number of binary
classifiers incorporated into the first layer have a great effect on
the final model, the general trend being that the more binary
classifiers are used, the better the nominal prediction.

@@

The classifier predicts whether a model has an RMSD of below 2~\AA,
between 2-4~\AA, or above 4~\AA. These cut-off values were selected
based on the observation that abYmod generally produces a model with
RMSD below 4~\AA. Incorrectly modelled structures (Figure~\ref{fig:AMA}) may
be identified by screening for structures estimated to have an RMSD
above 4~\AA. If a very high-quality model is needed one should also
exclude models above 2~\AA.

@@

\subsection{Computing}
All machine learning and, feature selection and hyperparameter
optimization algorithms were implemented in Python. The Scikit-learn
library was used for training models, the
Yellowbrick\cite{Bengfort2021} library was utilized for
visualization. All code is available at
https://github.com/LilianDenzler/qualiloop

@@

\subsubsection{Physiochemical Feature Encoding}
In a paper by L.~Nanni and A.~Lumini a new encoding technique is
presented which was developed for machine learning classifiers. Many
physiochemical properties are calculated and transformed using a
non-linear Fisher transform for dimensionality reduction.  A vector of
length 19 is produced for each amino acid34.  In a paper on designing
a neural network for predicting the packing angle of the light and
heavy variable chain of an antibodies, A.C.R.~Martin and
K.R.~Abhinandan introduce an encoding method that produces a
four-dimensional physiochemical feature vector25. The amino acid
properties used are 1) the total number of side-chain atoms, 2) the
number of side-chain atoms in the shortest path from \ca\ to the most
distal atom, 3) the Eisenberg consensus
hydrophobicity\cite{Eisenberg1982}, 4) the charge.

@@

\subsection{Model Assessment} 
Model assessment must be considered at two levels as performance
metrics of binary and multi-class classifiers are calculated
differently and must thus be considered separately. The Mathews
Correlation Coefficient (MCC)\cite{Matthews1975} is deemed the most
informative, taking the ratios of the four confusion matrix categories
into account 39and is thus more reliable than the F1 score and
accuracy. It is also consistent for both binary and multi-class
problems and therefore well suited for our purpose.

@@

\begin{figure}
  \includegraphics[scale=0.6]{angle.eps}
  \caption {Diagram visualizing the process underlying the protrusion
    calculation. First, the base residues (i.e.\ H95 and H102, shown as
    red spheres) of the CDR-H3 (shown in red) are identified. Then, a
    line is drawn between the two \ca\ atoms of these residues (yellow
    dashed line). The distance of the \ca-atom of each residue in the
    CDR-H3 loop to this line is calculated. The residue which has the
    greatest distance to the line (shown as green sphere) is outputted
    as one-letter amino acid code and used as feature. The distance in
    \AA\ (depicted as yellow arrow) is used as the `protrusion'
    feature.}
  \label{fig:angle}
\end{figure}

@@

Given that abYbank/AbDb is soon to be expanded by an additional
$\sim$2000 structures, classifier performance on models without a known
abYmod-template sequence may be improved by a larger dataset.
